Shaping strategies and standardisation of faceted spheroids from Qianshangying (North China)

### 2025-06-03 analysis/paper/paper.qmd
add some code blocks and inline code

We applied diacritical analysis to reconstruct the reduction sequence based on superimposition relationships among scars on spheroids [@Titton2020; @Cabanès2024]. This method enables the identification of scars order, allowing us to figure out the organization of removals. All spheroids, polyhedron and multifacial cores are included in the analysis to assess whether removals organization differs systematically within these groups.

@@

@@

We applied spherical sphericity index, harmonics analysis, and centroid offset measurements to both the spheroids (n = `r n_of_artefacts`) and all other multifacial and polyhedron cores (n = ) in the assemblage of QSY. These analyses aim to assess whether spheroids are morphologically distinct from other cores, and whether they exhibit standardized shapes—an essential criterion for evaluating the existence of deliberate shaping strategies and a goal product. In addition, we compare quantitative outcomes with traditional typological classifications to assess their consistency. We present a summary table listing each core’s prototype, initial and final shapes, and cortex coverage here.

Sphericity analysis reveals a clear difference between spheroids and multifacial cores/polyhedron. The mean sphericity of spheroids (0.906) is notably higher than that of multifacial cores and polyhedron (0.861), with a Mann-Whitney U test yielding U = 37.0, p = 3.99 × 10⁻⁵. This result indicates that spheroids possess a more spherical geometry overall, supporting their typological classification.

>>>>

We applied diacritical analysis to reconstruct the reduction sequence based on superimposition relationships among scars on spheroids [@tittonSubspheroidsLithicAssemblage2020; @Cabanes2024]. This method enables the identification of scars order, allowing us to figure out the organization of removals. All spheroids, polyhedron and multifacial cores are included in the analysis to assess whether removals organization differs systematically within these groups.

@@

```{r}
# multifacial and polyhedron cores
cores <- 
read_excel(here("analysis/data/derived_data/core technological parameters.xlsx"))

multifacial_and_polyhedron_cores <- 
    cores %>% 
  filter(!Typology %in% c("Spheroid", 
                          "Subspheroid", 
                          "Subsperoid"))

n_multifacial_and_polyhedron_cores <-
  nrow(multifacial_and_polyhedron_cores)
```

@@

```{r}
#| label: tbl-cores
#| tbl-cap: "Our table"

knitr::kable(multifacial_and_polyhedron_cores)
```

We applied spherical sphericity index, harmonics analysis, and centroid offset measurements to both the spheroids (n = `r n_of_artefacts`) and all other multifacial and polyhedron cores (n = `r n_multifacial_and_polyhedron_cores`) in the assemblage of QSY. These analyses aim to assess whether spheroids are morphologically distinct from other cores, and whether they exhibit standardized shapes—an essential criterion for evaluating the existence of deliberate shaping strategies and a goal product. In addition, we compare quantitative outcomes with traditional typological classifications to assess their consistency. We present a summary table listing each core's blank morphology and type, final morphology, and cortex coverage here in [@tbl-cores].

```{r}

# code for basic morphometry
mean_sphercity_tbl <- 
cores %>% 
  mutate(grouped_types = case_when(
    Typology == "Polyhedron" ~  "Multifacial",
    Typology == "Subspheroid" ~ "Spheroid",
    .default = Typology
  ))

# summary values for Sphericity
mean_sphercity_tbl_summary <- 
  mean_sphercity_tbl %>% 
  group_by(grouped_types) %>% 
  summarise(mean_sphercity = mean(Sphericity))

mean_sphercity_tbl_sph <- 
  mean_sphercity_tbl_summary %>% 
  filter(grouped_types == "Spheroid") %>% 
  pull(mean_sphercity) %>% 
  round(3)

mean_sphercity_tbl_mult <- 
  mean_sphercity_tbl_summary %>% 
  filter(grouped_types == "Multifacial") %>% 
  pull(mean_sphercity) %>% 
  round(3)

# Mann-Whitney U test
# Convert to scientific text
pretty_print_sci <- function(num){
  scientific_text <- paste0(gsub("e", " x 10^",  
                                 # Replace 'e' with ' x 10^'
                          sprintf("%.2e", num)), "^") 
  # round to 2 sf
  return(scientific_text)
}

# example usage:
# `r pretty_print_sci(journal_summary_metrics_ranks_test$p.value)
# convert 3.993e-5 to 3.99 × 10⁻⁵ for pretty printing in your paper

Sphericity analysis reveals a clear difference between spheroids and multifacial cores/polyhedra. The mean sphericity of spheroids (`r mean_sphercity_tbl_sph`) is notably higher than that of multifacial cores and polyhedron (`r mean_sphercity_tbl_mult`), with a Mann-Whitney U test yielding U = 37.0, p = 3.99 × 10⁻⁵. This result indicates that spheroids possess a more spherical geometry overall, supporting their typological classification.
